---
title: "Plots - BaseR"
author: "DSC130 - Exploring Data w/R"
date: '2022-09-12'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objectives

* Identify data types
* Plot categorical data
* Plot numerical data



---

<!----------------------------------------------------------------------------->
## Data Types

There are basically two data types:

1. Categorical (or nominal)
2. Numerical

The data type will dictate the plot function used.  








---

<!----------------------------------------------------------------------------->
## Categorical Data

Categorical data is associated with discrete variables (variables that can take 
finite values).  Categorical data is often referred to **group** variables.

#### Examples of Categorical data

1. Class rank: freshman, sophomore, junior, senior
2. Political party: Republican or Democrat
3. Blood type: A, B, AB, or O
4. Favorite Ice Cream: Chocolate or Vanilla
5. Nationality: French, German, Russian, ...
6. Parts of Speech: noun, verb, pronoun, preposition, ...

#### Plots for Categorical Data

1. Bar chart
2. Pie Chart
3. Violin Plot
4. Correlation Plot





---

<!----------------------------------------------------------------------------->
## Dataset

We will work with the `mtcars` data set bundled with R. 

```{r}
data(mtcars)
```


## Bar chart
```{r}
barplot(table(mtcars$cyl))

# See ?barplot
```
Alternatively, 
```{r}
# Alternative barplot method
plot(factor(mtcars$cyl))
```

### Grouped Barplot

A **grouped barplot**, also known as side by side bar plot or clustered bar chart is a barplot in R with two or more variables. The chart will display the bars for each of the multiple variables.

Look for common [color palettes](https://www.canva.com/colors/color-palettes/).


```{r}
# Short variable names. Note: Could "attach" instead
am = mtcars$am
cyl = mtcars$cyl

# Convert am to factor
am = factor(am)

# Change factor levels to readable names (instead 0/1)
levels(am) = c("Automatic", "Manual")

# Table cylinder v transmission type
cylvtrans = table(cyl, am)
# cylvtrans <- xtabs(~cyl + am , data = mtcars) # Equivalent

barplot(cylvtrans,
        main = "Grouped barchart",
        xlab = "Transmission type", ylab = "Frequency",
        col = c("#336699", "#80400B", "#FCAE1E"),
        legend.text = rownames(cylvtrans),
        beside = TRUE) # Grouped bars
```
### Stacked Barplots

A stacked bar chart is like a grouped bar graph, but the frequency of the variables are stacked. This type of barplot will be created by default when passing as argument a table with two or more variables, as the argument beside defaults to FALSE.

```{r}
barplot(cylvtrans,
        main = "Stacked barchart",
        xlab = "Transmission type", ylab = "Frequency",
        col = c("#FF8370", "#00B1B0", "#FEC84D"),
        legend.text = rownames(cylvtrans),
        beside = FALSE) # Stacked bars (default)
```
### Mosaic plots

Search internet for mosaic plots in R.  You will need to install a library called: "graphics".



---

<!----------------------------------------------------------------------------->
### Pie Chart

A pie chart, also known as circle chart or pie plot, is a circular graph that represents proportions or percentages in slices, where the area and arc length of each slice is proportional to the represented quantity. Variations of this type of chart are doughnut charts, waffle charts and pie chart.


The R pie function allows you to create a pie chart in R. Consider, for instance, that you want to create a piechart of the following variable, that represents the count of some event:

```{r}
pie.sales = c(0.12, 0.3, 0.26, 0.16, 0.04, 0.12)
names(pie.sales) = c("Blueberry", "Cherry",
    "Apple", "Boston Cream", "Other", "Vanilla Cream")
```


The code for a pie chart in R is as follows. Note that you can customize the size of the pie (from -1 to 1) with the radius argument, that by default takes the value 0.8.

```{r}
pie(pie.sales, col=rainbow(6))

# See ?pie
```

---

<!----------------------------------------------------------------------------->
## Histogram (numerical data)

Displays frequency (or relative frequency) distribution.  

```{r}
hist(mtcars$mpg)

# See ?hist
```
```{r}
hist(mtcars$mpg, freq = FALSE, main = "Histogram and density")

# Calculate density
dx = density(mtcars$mpg)

# Add density
lines(dx, lwd = 2, col = "red")


```
### Boxplots 

A **boxplot** in R, also known as box and whisker plot, is a graphical representation which allows you to summarize the main characteristics of the data (position, dispersion, skewness, etc.) and identify the presence of outliers.

The box of a boxplot starts in the first quartile (25%) and ends in the third (75%). Hence, the box represents the 50% of the central data, with a line inside that represents the median. On each side of the box there is drawn a segment to the furthest data without counting boxplot outliers, that in case there exist, will be represented with circles.



```{r}
boxplot(mtcars$mpg)
```

Data with outliers:

```{r}
x = c(8, 5, 14, -9, 19, 12, 3, 9, 7, 4, 4, 6, 8, 12, -8, 2, 0, -1, 5, 3)
boxplot(x, horizontal = TRUE, col=2)
```
Display data too.

```{r}
boxplot(x, horizontal = TRUE, col=2)
stripchart(x, method = "jitter", pch = 19, add = TRUE, col = "blue")
```
#### Multiple Boxplots

```{r}
boxplot(wt ~ cyl, data = mtcars, col = rainbow(length(unique(mtcars$cyl))))
```
```{r}
boxplot(wt ~ cyl, data = mtcars, col = rainbow(length(unique(mtcars$cyl))))
stripchart(mtcars$wt ~ mtcars$cyl, vertical = TRUE, method = "jitter",
           pch = 19, add = TRUE, col = 4:4+length(levels(mtcars$cyl)))
```

---

<!----------------------------------------------------------------------------->
### Scatter plot

Scatter plots are used for **bivariate** data (e.g., height vs. weight).  Displays
relationship between two variables.  

```{r}
# mpg vs. displacement
plot(mtcars$disp,mtcars$mpg)
```
```{r}
sort(c(unique(mtcars$cyl)))
```


```{r}
plot(mtcars$disp,mtcars$mpg, col=mtcars$cyl)  # 3rd variable color

# Add legend to plot
legend("topright",                    
       legend = c("4", "6", "8"),
       col=sort(c(unique(mtcars$cyl))),
       pch = 20)

```


```{r}
cols = c("deepskyblue", "magenta", "gray50")
plot(mpg ~ wt, xlab="Weight (1000lbs)", ylab="Miles / gallon", col = cyl, pch = 19, data=mtcars)
legend(x = "topright",
       title = "Number of cylinders",
       horiz = TRUE,
       legend = c("4", "6", "8"), 
       bty = 'n',
       cex = 0.7,
       col = cols,
       pch=20)

```

---

<!----------------------------------------------------------------------------->
### Plot Attributes

You can control many features of a plot (e.g., color, title, x-axis labels, etc.).

```{r}
plot(mtcars$disp,mtcars$mpg, 
     main = "MPG vs. Displacement", 
     xlab = "Displacement",
     ylab = "MPG", 
     col = "#336699", 
     pch = 19)

# See ?plot
```
```{r}
plot(mtcars$disp,mtcars$mpg, col=2)
grid(nx = NULL, ny = NULL,
     lty = 2,      # Grid line type
     col = "gray", # Grid line color
     lwd = 2)      # Grid line width
```
```{r}
hist(mtcars$mpg)
grid(nx = NA, ny = NULL,
     lty = 1, col = "gray", lwd = 1)
# hist(mtcars$mpg, col = 4, add = TRUE)
```

### Backgrounds

```{r}
# Data
set.seed(132)
x <- rnorm(500)

boxplot(x)

# Light gray background
rect(par("usr")[1], par("usr")[3],
     par("usr")[2], par("usr")[4],
     col = "#ebebeb")
# See: https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/par


# Add white grid
grid(nx = NULL, ny = NULL, col = "white", lwd = 2)

# Boxplot
boxplot(x, col = 4, add = TRUE)
```


### Multiple Plots

<!-- https://r-charts.com/base-r/combining-plots/ -->

```{r}
# Data
x = rnorm(50)

# One row, two columns
par(mfrow = c(1, 2))

# The following two plots will be combined
hist(x, main = "Left plot")     # Left 
boxplot(x, main = "Right plot") # Right

# RESET to the original graphics device
par(mfcol = c(1, 1)) 

```

```{r}
# Data
set.seed(6)
x = rexp(50)

# Two rows, two columns
par(mfcol = c(2, 2))

# Plots
hist(x, main = "Top left")                   # Top left
boxplot(x, main = "Bottom left")             # Bottom left
plot(x, main = "Top right")                  # Top right
pie(table(round(x)), main = "Bottom right")  # Bottom right

# RESET to the original graphics device
par(mfcol = c(1, 1)) 

```


### Annotation

```{r}
# Data
set.seed(123)
x = 1:50
y = x + rnorm(50, sd = 10)
groups = sample(letters, 50, replace = TRUE)
```

```{r}
plot(x, y, pch = 19, col = 4)
text(x = 15, y = 50, label = "Text annotation",
     col = "blue",   # Color of the text
     font = 2,      # Bold face
     cex = 1.1)     # Size 
```


```{r}
plot(x, y, pch = 19, col = 4)

# Split the text in several lines
text(x = 15, y = 50,
     label = "Text\n annotation") # Split text 

```

```{r}
plot(x, y, pch = 19, col = 4,
     ylim = c(-10, 70)) # Change axis limits
text(x = x, y = y - 3,  # Fine-tune the position
     label = groups)  

# Config axis
# Add X-axis
axis(1,
     col = "blue",        # Axis line color
     col.ticks = "green", # Ticks color
     col.axis = "red")    # Labels color

```


### Muliple data one graph

Here's some data
```{r}
# Reproducible data
set.seed(1)
x = runif(100, min = -5, max = 5)
y = x ^ 3 + rnorm(100, mean = 0, sd = 5)
```


```{r}
# Increase the plot margins
par(mar = c(5, 4, 4, 4) + 0.25)

# First data plot
plot(x, y, pch = 19, ylab = "Var 1")

# SET TRUE to merge plots
par(new = TRUE)

# Another data set to plot
plot(x ^ 2, y ^ 2,
     col = 4, pch = 19,
     axes = FALSE, # No axes
     bty = "n",    # No box
     xlab = "", ylab = "") # No axis labels

# New axis on right with proper scale
axis(4)

# Axis label (if needed)
mtext("Var 2", side = 4, line = 3, col = 4) 
```