---
title: "Killings by Police"
author: "DSC130"
date: "Assignment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<style>
p{font-size:15px;}
h1{font-weight: bold;}
h2{color: #336699;}
h3{color: #666666;}
ul{font-size:14px;}
ol{font-size:15px;}
</style>


# Assignment Tasks

##### 1. Obtain the raw data from the source (see # Sources)

* Can you trust this data set?  

* How accurate is the data?  

* What is the source?

> Any analysis, results, and conclusions are only as good as your data.

```{r}
killings = read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/police-killings/police_killings.csv')
```


```{r}
killings$age = as.integer(killings$age)
```

```{r}
mean(killings$age, na.rm = TRUE)
```


##### 2. Determine how many records the data set contain. 

```{r}
# [1] Rows = observations = records
```

##### 3. How many variables? 

```{r}
#
```

##### 4. How many different data types are there? 

```{r}
#
```

##### 5. What are the data types? (Make a list)  
```{r}
#
```
**Answer**: 

-  
- 

---


##### 6. What type is age? Why?  What type should it be?

**Answer**:  

```{r}
#
```





##### 7. What is the average age of deceased? Round to nearest integer.

```{r}
#
```


#### 8. Which state had most police killings?

```{r}
#
```



##### 9. What is the youngest person killed by police? 

```{r}
#
```




##### 10. How many deceased were armed with a knife?
```{r}
#
```


##### 11. How many died as a result of being tased?
```{r}
#
```

##### 12. Compare deaths by race/ethnicity. Use a barchart.


```{r}
#
```

##### 13. List all names and the city/state of the deceased (order by city/state).

```{r}
#
```




### Create your own questions (at least 5) and provide answers.

1. 

1. 

1. 

1.

1.



---






 


##  Sources


[Data](https://raw.githubusercontent.com/fivethirtyeight/data/master/police-killings/police_killings.csv)

[Data Definitions](https://github.com/fivethirtyeight/data/tree/master/police-killings)

[Wikipedia Listings](https://en.wikipedia.org/wiki/List_of_unarmed_African_Americans_killed_by_law_enforcement_officers_in_the_United_States#2015)





